Longitudinal Principal Manifold Estimation

### 2024-03-02 paper/main.tex

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We then apply the proposed method to estimating the surfaces of hippocampi of healthy individuals in the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{high-dimensional data tend to lie in the vicinity of a low-dimensional manifold} (\cite{fefferman2016testing}). Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum. To obtain the low-dimensional parameterization of the substructures of interest, the proposed approach extends the principal manifold estimation approach introduced by \cite{mengPrincipalManifoldEstimation2021}.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we use a process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

The remainder of this article is laid out as follows: In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus and thalamus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}). Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing). At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping. Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space. For example, the PME algorithm can be used to obtain this parameterization. In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, smoothing over the spatial spline coefficients with respect to time. This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. A visualization of these steps is shown in Figure \ref{fig:lpme_steps}. Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \texttt{https://github.com/rjzielinski/pme}.

>>>>

Alzheimer's disease (AD) is a neurogenerative disorder affecting, among others, the structure of the brain. Longitudinal magnetic resonance imaging (MRI) data is used to model trajectories of change in brain regions of interest to identify areas more susceptible to atrophy. Most methods for extracting surfaces of brain regions are applied to individual scans from study participants independently. As a result, there is a wide variability of shape and volume estimates of brain regions of interest over time in longitudinal studies resulting in major implications for biomarker estimation and modeling, especially when used in therapeutic clinical trials. To address this problem, we propose a longitudinal principal manifold estimation method, with the goal of recovering smooth, longitudinally meaningful manifold estimates of shapes over time. The proposed approach uses a smoothing spline to smooth over the coefficients of the principal manifold embedding function estimated at each time point. This smoothing mitigates the effects of random disturbances to the manifold between time points. A novel data augmentation approach is used to allow the use of principal manifold estimation on self-intersecting manifolds. We use simulation studies with several classes of manifolds to demonstrate performance improvements over naïve applications of principal manifold estimation and principal curve/surface methods. These improvements persist when considering varying between-time-point noise levels and the types and magnitudes of systematic change between time points. We show that the proposed method improves the estimation of surfaces of hippocampi and thalamuses of healthy individuals using data from the Alzheimer’s Disease Neuroimaging Initiative dataset.

@@

Manifold learning describes a set of statistical approaches to modeling high-dimensional data that satisfy the ``manifold hypothesis" --- \textit{``high-dimensional data tend to lie in the vicinity of a low-dimensional manifold''} (\cite{fefferman2016testing}). Manifold learning approaches have found applications in various scientific fields, e.g., medical signal processing (\cite{shen2022robust}) and robotics (\cite{gao2023k}). In medical imaging applications, interacting with a low-dimensional interpretation of a structure may be more intuitive and efficient than working with the structure in its original high-dimensional space. For example, \cite{yueParameterizationWhiteMatter2016} seeks to compute a parametric representation of the corpus callosum that is used for investigating the features of this structure in people with multiple sclerosis.

Previously introduced methods for manifold learning, such as Isomap (\cite{tenenbaumGlobalGeometricFramework2000}), locally linear embedding (\cite{roweisNonlinearDimensionalityReduction2000}; \cite{wu2018think}), and Laplacian eigenmaps (\cite{belkin2003laplacian}), are ill-equipped to reach meaningful estimates across multiple time points. These methods consider time as an additional dimension in the data, i.e. they include time in the dimension reduction process. This treatment of the time dimension causes it to be incorporated into the low-dimensional parameterization, preventing meaningful interpretation of this dimension. Here, we propose an alternative approach to estimating manifolds over time. Rather than include time as another dimension in a manifold learning method, we introduce a novel process that maintains the interpretability of the time dimension. Specifically, we use the method proposed in \cite{mengPrincipalManifoldEstimation2021} to model the appropriate principal manifold at each given time point. We then build on this model by imposing smoothness over these approximated manifolds in the time dimension, yielding an estimate showing longitudinal changes in the underlying manifold. We demonstrate that the regularization involved in this process mitigates the effects of variability between study visits.

@@

The remainder of this article is laid out as follows. In Section \ref{s:PME}, we review the principal manifold framework. In Section \ref{s:LPME} we discuss our approach to adapting the principal manifold framework for longitudinal settings, and show the algorithm we use to accomplish dimension reduction using the longitudinal principal manifold framework. Section \ref{s:simulations} demonstrates the performance of this approach on simulated data. In Section \ref{s:application}, we apply our proposed method to longitudinal data of the hippocampus and thalamus in individuals with Alzheimer's disease. The article concludes with a discussion of the method's contributions in Section \ref{s:discussion}.

@@

Given the framework proposed in the previous section, estimation of the longitudinal principal manifold entails minimizing the function in (\ref{eq:newKappa}). Our proposed algorithm is based on a multi-stage approach, where we consider smoothing of the data clouds at each time point (spatial smoothing) and then smoothing of the obtained manifolds over time (temporal smoothing). At each individual time point we assume that a manifold has been fit to the data cloud using a spline-based mapping. Hence, in the spatial smoothing phase we assume that we have a collection of spline coefficients that can be used to project from the low-dimensional parameterization to the high-dimensional space. For example, the PME algorithm can be used to obtain this parameterization. In the LPME algorithm, we then use these low-dimensional representations at each time point as the outcomes in a thin plate spline, smoothing over the spatial spline coefficients with respect to time. This yields a multi-stage model that, for a provided time point and set of tuning parameters, obtains a set of spline coefficients that are time-dependent, then returns the high-dimensional output of the spline function associated with the tuning parameter set. Our proposed LPME algorithm consists of four steps: data reduction, initialization, fitting, and tuning. A visualization of these steps is shown in Figure \ref{fig:lpme_steps}. Details of this approach are described below and formally given in Algorithm \ref{alg:lpme}, and an implementation is available as an \texttt{R} package at \href{https://github.com/rjzielinski/pme}{\texttt{https://github.com/rjzielinski/pme}}.
